---
title: "Cubic Zirconia"
author: "Ally Weingarden, Juliet Niebylski, Ziying Peng, Leah Sun, Marisa Wong"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data from: https://www.kaggle.com/datasets/colearninglounge/gemstone-price-prediction?select=cubic_zirconia.csv

For the second group project, we have chosen a dataset on cubic zirconia. This dataset has 10 variables which includes the price as well as various characteristics about the cubic zirconia. In our project, we would predict the price of a cubic zirconia, given certain characteristics. This addresses the business problem of setting prices for a seller's inventory of cubic zirconia since it is important to set reasonable prices based on what other similar cubic zirconia are selling for.

```{r}
cubicz <- read.csv("cubic_zirconia.csv")
```

```{r}
str(cubicz)
```

# Clean Data

```{r}
cubicz$X <- NULL
cubicz$cut <- as.factor(cubicz$cut)
cubicz$color <- as.factor(cubicz$color)
cubicz$clarity <- as.factor(cubicz$clarity)
#cubicz$depth <- ifelse(is.na(cubicz$depth) == TRUE, mean(cubicz$depth),cubicz$depth)
#cubic <- tapply(ifelse(is.na(cubicz$depth) == TRUE,mean,na.rm=T))
str(cubicz)
summary(cubicz)
summary(cubicz$depth)
cubicz$depth <- ifelse((is.na(cubicz$depth) == TRUE), mean(cubicz$depth), cubicz$depth)
summary(cubicz$depth)
```

Before moving to the analysis of the data set, we cleaned the data to prepare it for regression. The variablee of X is an index variable, which is not necessary for our analysis. Cut, Color, and Clarity all have a clear number of levels, so we converted those variables into factors. 

# Regression

```{r}
cubicz_model <- lm(price ~ ., data = cubicz)
summary(cubicz_model)
```

